---
title: "01_Mars_Shape_Phylo_Prep"
author: "Vera Weisbecker"
date: "26 June 2019"
output: html_document
---


#Set working directory, load required packages

```{r }

#This line sets the working directory to be where this code is stored. 
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


## Loading the libraries (and installing if necessary)
if(!require(geomorph)) install.packages("geomorph")
if(!require(abind)) install.packages("abind")

```

# Load 3d coordinates from the files you exported from Checkpoint
 
```{r }

filelist <- list.files(path= "../Data/Raw/Coordinates", pattern = "*.txt")

#Next step is to remove the Museum IDs from the specimen names   
  
  tmp <- matrix(unlist(strsplit(filelist, "_")),ncol=3,byrow = TRUE)   

  filelist_species <- paste(tmp[,1], tmp[,2], sep="_")

  
  names <- gsub (".txt", "", filelist) # extracts names of specimens from the file name
  filelist <- paste("../Data/Raw/Coordinates/", filelist, sep="") # rename with path
  coords <- NULL # make empty object that will be filled with 3D array of coordinate data
  for (i in 1:length(filelist)){
    temp  <- read.morphologika(filelist[i]) 
    k <- dim(temp)[1] 
    coords <- rbind(coords, two.d.array(temp)) }
  Data <- arrayspecs(coords, k, 3) 
  dimnames(Data)[[3]] <- filelist_species
  remove(i, filelist, names, k, coords, temp) # clean up environment
  
  #Check the data read in ok. $dim should have number of landmarks (58), number of dimensions (3), number of specimens (10), and teh specimen names.
  
  attributes(Data)
  
  #To navigate to a particular specimen (e.g. the second one in the array, which you know from the attributes() command above is called "Aepyprymnus_rufescens")
  Data[,,2]
  
  
```


# You can  read in a surface file if you want - this is useful for later visualising shape change and making sure the coords and the 3D files line up. You might want to do this later if you want to warp a mesh to the mean shape (NOT explained here, just saying)

```{r}


#This is a mesh I crunched down to 600,000 faces. It's 18 kb.
Apyg <- read.ply("../Data/Raw/Acrobates_pygmaeus_J7579.ply")

shade3d(Apyg)

#There's a weird thing about these meshes at the moment, they show up as black. So to fix this I found a strange workaround - but any improvements are much appreciated

Apyg$material$color <- Apyg$it
Apyg$material$color <- "hotpink"

#Isn't she gorgeous!!
shade3d(Apyg)


```


#If you have curve or patch semilandmarks, you need to make a separate file for them that allows them to be identified. I won't do it today but the best way to do this is to plot the landmarks: 

```{r}
#Double check that the 3D info is entered properly by checking landmark numbers
  plot3d(Data[,,4], asp=FALSE, col="hotpink")
  text3d(Data[,,4], texts=c(1:58))
  

#open new 3D window so any previous don't accidentally superimpose  
open3d()

#There's a way of getting rid of those lines but I forgot it. Note that you MUST specify at least the fixed landmark number to see any landmarks. It won't just automatically show all landmarks.
plotspec(Apyg, Data[,,1], fixed=58, mesh.ptsize = 0.00001 )
text3d(Data[,,1], texts=c(1:58))
```


#OR:

```{r}
open3d()

#a transparent 3d object. alpha is 0=transparent, 1=opaque
shade3d(Apyg, alpha = 0.4)
# put on the landmarks as spheres, radius is how big
spheres3d(Data[,,1],  col="green", radius = 0.1)
#text for landmark numbers, e.g. if you want to determine the lm numbers for your patches and curves
text3d(Data[,,1], texts=c(1:58), pos = 2, cex=1.2, font = 2)

```

# if you are using patch/curve semilandmarks, use the above to figure out what numbers they are. I normally enter them into csv files manually. I have example csv files in teh Data/Raw folder that doesn't match this dataset,just fyi. You read them in like:  

```{r}

#curveLM <- read.csv(curveslides)

#surfaceLM <- read.csv(surfslides)


```


#Read a classifier file that contains the species name, and info on locomotion type, diet, and clade; this classifier file is used for analysing shape relative to whatever data are entered in that file for all species. 


```{r}

speclist<- read.delim("../Data/Raw/Trial_classifier_list.txt")
rownames(speclist) <-speclist$Specimen

#Double check that the order of your Array and the order of your specimen list are the same (they will need to talk to each other!). This has to return "TRUE" for all. 

dimnames(Data)[[3]]==rownames(speclist)

```


#GPA for all coordinates

```{r}
#Run GPA
GPA <- gpagen(Data)

attributes(GPA)

GPA$coords[,,1]

#Find the mean shape

findMeanSpec(GPA$coords)

Ref <- mshape(GPA$coords)



```
#run a basic PCA
```{r}

PCA <- gm.prcomp(GPA$coords)

attributes(PCA)

plotRefToTarget(Ref, PCA$shapes$shapes.comp1$max, method="vector")

plot(PCA$x[,1]~ log(GPA$Csize))


```



#Package into RDA file. This is useful if you want to keep everything your computer just did and re-open it for use in another file. For example, you might use this file to do the basic loading, GPA, PCA of data but then later do something more fancy or a specific analysis with everything already loaded. 
```{r}

save.image(file="../Data/Processed/Read_in_Data.rda")

#Now if you clear the environment
rm(list = ls())


#You can now get it all back by doing:

load (file="../Data/Processed/Read_in_Data.rda")


```

