---
title: "01_Mars_Shape_Phylo_Prep"
author: "Vera Weisbecker"
date: "26 June 2019"
output: html_document
---


#Set working directory, load required packages

```{r }

#This line sets the working directory to be where this code is stored. At least on most computers. If this doesn't work, you should be able to just enter the whole file path wherever you need to read a file (or write to a file). In that case, beware of the slashes - R works with forward slashes (/) but PCs tend to work with backward slashes. There is a way of forcing slashes the right way but it's easier to just manually change.
library(rstudioapi)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))


## Loading the libraries (and installing if necessary)
if(!require(geomorph)) install.packages("geomorph")
if(!require(abind)) install.packages("abind")

```

# Load 3d coordinates from the files you exported from Checkpoint. This is for Morphologika files and the code comes from Emma Sherratt
 
```{r }

filelist <- list.files(path= "../Data/Raw/Coordinates", pattern = "*.txt")

#Next step is to remove the Museum IDs from the specimen names (in this particular case. Each dataset is different but each specimen needs a unique name and the specimen names need to match up with any classifier files you are using)  
  
  tmp <- matrix(unlist(strsplit(filelist, "_")),ncol=3,byrow = TRUE)   

  filelist_species <- paste(tmp[,1], tmp[,2], sep="_")

  
  names <- gsub (".txt", "", filelist) # extracts names of specimens from the file name
  filelist <- paste("../Data/Raw/Coordinates/", filelist, sep="") # rename with path
  coords <- NULL # make empty object that will be filled with 3D array of coordinate data
  
  for (i in 1:length(filelist)){
    temp  <- read.morphologika(filelist[i]) 
    k <- dim(temp)[1] 
    coords <- rbind(coords, two.d.array(temp)) }
  Data <- arrayspecs(coords, k, 3) 
  dimnames(Data)[[3]] <- filelist_species
  remove(i, filelist, names, k, coords, temp) # clean up environment
  
  #Check the data read in ok. $dim should have number of landmarks (58), number of dimensions (3), number of specimens (10), and the specimen names.
  
  attributes(Data)
  
  #To navigate to a particular specimen (e.g. the second one in the array, which you know from the attributes() command above is called "Aepyprymnus_rufescens")
  Data[,,2]
  
  
  #Double check that the 3D info is entered properly by checking landmark numbers
  plot3d(Data[,,4], asp=FALSE, col="hotpink")
  text3d(Data[,,4], texts=c(1:58))
  
```


# You can  read in a surface file if you want - this is useful for later visualising shape change and making sure the coords and the 3D files line up. You might want to do this later if you want to warp a mesh to the mean shape (NOT explained here, just saying)

```{r}

#This is a mesh I crunched down to 600,000 faces. It's 18 kb.
Apyg <- read.ply("../Data/Raw/Acrobates_pygmaeus_J7579.ply")

shade3d(Apyg)

#There's a weird thing about these meshes at the moment, they show up as black. So to fix this I found a strange workaround - but any improvements are much appreciated

Apyg$material$color <- Apyg$it
Apyg$material$color <- "hotpink"

#Isn't she gorgeous!!
shade3d(Apyg)


```


#If you have curve or patch semilandmarks, you need to make a separate file for them that allows them to be identified. I won't do it here but the best way to determine which curve landmarks are which number is to plot the landmarks: 

```{r}

  
open3d()

#a transparent 3d object. alpha is 0=transparent, 1=opaque
shade3d(Apyg, alpha = 0.4)
# put on the landmarks as spheres, radius is how big
spheres3d(Data[,,1],  col="green", radius = 0.1)
#text for landmark numbers, e.g. if you want to determine the lm numbers for your patches and curves
text3d(Data[,,1], texts=c(1:58), pos = 2, cex=1.2, font = 2)

```

# if you are using patch/curve semilandmarks, use the above to figure out what numbers they are. I normally enter them into csv files manually. I have example csv files in teh Data/Raw folder that doesn't match this dataset,just fyi. You read them in like:  

```{r}

#curveLM <- read.csv(curveslides)

#surfaceLM <- read.csv(surfslides)


```


#Read a classifier file that contains the species name, and info on locomotion type, diet, and clade; this classifier file is used for analysing shape relative to whatever data are entered in that file for all species. 


```{r}

speclist<- read.delim("../Data/Raw/Trial_classifier_list.txt")
rownames(speclist) <-speclist$Specimen

#Double check that the order of your Array and the order of your specimen list are the same (they will need to talk to each other!). This has to return "TRUE" for all. 

dimnames(Data)[[3]]==rownames(speclist)

```


#GPA for all coordinates

```{r}
#Run GPA
GPA <- gpagen(Data)

#This is waht's in a GPA. Note the separate Centroid sizes - a good way of checking if they are all correct is to check if your smallest specimen also has the smallest C size. There is a lot of potential for error in specimen dimensions so it's worth keeping a good eye on the Csizes at all times. 

attributes(GPA)

GPA$coords[,,1]

#Find the specimen that is closest to the overall mean shape

findMeanSpec(GPA$coords)

#Or create a reference configuration taht is the actual mathematical mean shape:

Ref <- mshape(GPA$coords)



```

#run a basic PCA, which arranges teh specimens according to where the greates amount of variation is (PC1), then the 2nd greatest (PC2)... etc. This is good for summarising but don't be tempted to analyse the PCAs much. There are better ways of asking if there is a difference between groups than checking a PCA plot.

```{r}

PCA <- gm.prcomp(GPA$coords)

#This is what's in a PCA
attributes(PCA)

#How are specimesn arranged on a PC1 vs PC 2 plot?


plot(PCA$x[,2]~ PCA$x[,1], pch=19)
text(  PCA$x[,2]~ PCA$x[,1], labels=rownames(speclist)  )

# How does PC1 relate to centroid size? for fiddling with image settings, go to R console and type in ?plot. Note I'm log-transforming csize because animals tend to evolve on an exponential scale when it comes to size. If we didn't log, we would have all the small specimens bunched on the left and the relationship would likely not be linear.

plot(PCA$x[,1]~ log(GPA$Csize), pch=19)
text(  PCA$x[,2]~ PCA$x[,1], labels=rownames(speclist)  )

#This shows the differences between a hypothetical specimen's landmark configuration at the lower scores of PC1 - balls - and compares it to a hypothetical landmark configuration for the higher scores of PC1 (where the hairs point to)
plotRefToTarget(PCA$shapes$shapes.comp1$min, PCA$shapes$shapes.comp1$max, method="vector")




```



#Package into RDA file. This is useful if you want to keep everything your computer just did, and re-open it for use in another file. For example, you might use this file to do the basic loading, GPA, PCA of data but then later do something more fancy or a specific analysis with everything already loaded. 
```{r}

save.image(file="../Data/Processed/Read_in_Data.rda")

#Now if you clear the environment
rm(list = ls())

#You can now get it all back by doing:

load (file="../Data/Processed/Read_in_Data.rda")


```

